Constructing a universe for the setoid model

### 2021-01-21 paper.tex
small changes
after meeting
Clarified sort hierarchy. @Ambrus: Check for consistency with formalization.

One of the main appeals of Altenkirch's setoid model is that it can justify
several useful extensionality principles while being defined in a small
intensional metatheory.
%
We tried to stay true to this idea when figuring out the necessary
metatheoretical tools for the universe construction in this paper. In
particular, we wanted to avoid having to assume strong definition schemas that
go beyond inductive families.

$\mlttp$ is thus an intensional type theory in the style of Martin-L\"of type
theory with $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$, and a universe
of strict propositions $\mProp$. $\Sigma$ types are defined negatively by
pairing $\_,\_$ and projections $\pi_1, \pi_2$. We also have definitional $\eta$
rules for $\Pi, \Sigma$, and $\mathbf{1}$ types. Propositions can be lifted to
types via a lifting $\textsf{Lift} : \mProp \to \mType$, with constructor
$\textsf{lift} : \{P : \mProp \} \to P \to \textsf{Lift}\ P$ and destructor
$\textsf{unlift} : \{P : \mProp \} \to \textsf{Lift}\ P \to P$.

We also require indexed W-types, both in $\mType$ and $\mProp$; hence we have $W_\Box
: (S : I\to\mType) \to ((i:I)\to S\ i\to I \to \mType) \to I\to \Box$ where $\Box \in \{\mType, \mProp\}$.
The elimination principle of $W_\mProp$ only allows defining functions into $\mProp$s.
%
From $W_\mProp$ we can define propositional truncation $\|\_\| : \mType \to \mProp$, with constructor
  $|\_| : \{A : \mType \} \to A \to \| A \|$ and eliminator
  $\textsf{elim}_{\|\_\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P$.

@@

Induction-recursion is arguably a nice and natural way to define internal
universes in type theory, however it is not always strictly required.
%
We can translate basic instances of induction-recursion into inductive families
using the equivalence of $I$-indexed families of types and types over $I$
(that is, $\mType_i \to I$) \cite{malatasta13smallir}.

>>>>

One of the main appeals of Altenkirch's setoid model is that it can justify
several useful extensionality principles while being defined in a small
intensional metatheory.
%
We tried to stay true to this idea when figuring out the necessary
metatheoretical tools for the universe construction in this paper. In
particular, we wanted to avoid having to assume strong definition schemas that
go beyond inductive families.
$\mlttp$ is thus an intensional type theory in the style of Martin-L\"of type theory.

We have sorts $\Type_i$ of types and $\mProp_i$ of strict propositions for $i \in \{0, 1\}$.
Here, $i = 0$ means ``small'' (and we will omit the subscript) and $i = 1$ means ``large''.
We have implicit lifting from $i = 0$ to $i = 1$, but do not assume type formers are preserved.
$\Type_1$ has universes for $\Type$ and $\mProp$.
We do not distinguish notationally between universes and sorts.
We continue to describe only the case $i = 0$; everything introduced has an analogue at level $i = 1$.
Propositions lift to types via $\textsf{Lift} : \mProp \to \mType$, with constructor
$\textsf{lift} : \{P : \mProp \} \to P \to \textsf{Lift}\ P$ and destructor
$\textsf{unlift} : \{P : \mProp \} \to \textsf{Lift}\ P \to P$.

We have standard type formers $\Pi, \Sigma, \textsf{Bool}, \mathbf{0}, \mathbf{1}$ in $\Type$.
$\Sigma$-types are defined negatively by pairing $\_,\_$ and projections $\pi_1, \pi_2$.
We also have definitional $\eta$-rules for $\Pi, \Sigma$, and $\mathbf{1}$ types.
We also require indexed W-types, both in $\mType$ and $\mProp$; hence we have $W_\Box
: (S : I\to\mType) \to ((i:I)\to S\ i\to I \to \mType) \to I\to \Box$ where $\Box \in \{\mType, \mProp\}$.
The elimination principle of $W_{\mProp}$ only allows defining functions into elements of $\mProp$.
%
From $W_{\mProp}$ we can define propositional truncation $\|\_\| : \mType \to \mProp$, with constructor
  $|\_| : \{A : \mType \} \to A \to \| A \|$ and eliminator
  $\textsf{elim}_{\|\_\|} : \{P : \mProp \} \to (A \to P) \to \| A \| \to P$.

@@

Induction-recursion is arguably a nice and natural way to define internal
universes in type theory, however it is not always strictly required.
%
We can translate basic instances of induction-recursion into inductive families
using the equivalence of $I$-indexed families of types and types over $I$
(that is, $\mType_i \to I$) \cite{malatasta13smallir}. % TODO: fix wrong formula
